---
title: "Political Text Messaging Analysis"
author: "Divya Parmar"
date: "April 20, 2018"
output: 
  html_document:
    keep_md: true
---

###Setup

Clear workspace before starting analysis.

```{r}
rm(list = ls())
```

Install necessary packages.

```{r results='hide', warning=FALSE, message = FALSE, eval=FALSE}

install.packages('dplyr')
library(dplyr)
library(ggplot2)
install.packages('plotrix')

```

Create t-test function to use later on.

```{r}

t.test2 <- function(m1,m2,s1,s2,n1,n2,m0=0,equal.variance=FALSE)
{
  if( equal.variance==FALSE ) 
  {
    se <- sqrt( (s1^2/n1) + (s2^2/n2) )
    # welch-satterthwaite df
    df <- ( (s1^2/n1 + s2^2/n2)^2 )/( (s1^2/n1)^2/(n1-1) + (s2^2/n2)^2/(n2-1) )
  } else
  {
    # pooled standard deviation, scaled by the sample sizes
    se <- sqrt( (1/n1 + 1/n2) * ((n1-1)*s1^2 + (n2-1)*s2^2)/(n1+n2-2) ) 
    df <- n1+n2-2
  }      
  t <- (m1-m2-m0)/se 
  dat <- c(m1-m2, se, t, 2*pt(-abs(t),df))    
  names(dat) <- c("Difference of means", "Std Error", "t", "p-value")
  return(dat) 
}

```

###Exploration

Read in data files.


```{r}
rand_data <- read.csv("C:\\Users\\Divya Parmar's PC\\Documents\\Task_Data_Analyst_Institute\\data\\randomization.csv")
sms_data <- read.csv("C:\\Users\\Divya Parmar's PC\\Documents\\Task_Data_Analyst_Institute\\data\\text_message_data.csv")
survey_data <- read.csv("C:\\Users\\Divya Parmar's PC\\Documents\\Task_Data_Analyst_Institute\\data\\survey_data.csv")
turnout_data <- read.csv("C:\\Users\\Divya Parmar's PC\\Documents\\Task_Data_Analyst_Institute\\data\\turnout_data.csv")
```

Look at data to understand columns.

```{r}

head(rand_data)
head(sms_data)
head(survey_data)
head(turnout_data)

```

Look at control/treatment breakdown for entire universe of voters.

* Treatment has 12,461 individuals

* Control has 12,539 individuals

* It looks like an evenly balanced A/B test.

```{r}

table(rand_data$sms_treat)

```

Look at texting data to make sure experiment was executed correctly.

* There are 12,461 voters in the treatment group.
* All of the phone numbers got a voting reminder.
* All of the phone numbers got a environment text message.
* This means the treatment is working as intended.

```{r}
#How many unique phone numbers got sent text messages -> 12,461
outbound_messages = sms_data[sms_data$message_direction == 'outbound',]
length(unique(outbound_messages$phone_number))

#How many unique phone numbers got the voting reminder text message -> 12,461
#We can conclude that all individuals in the treatment group got reminder text messages as intended
reminder_messages = sms_data[sms_data$message_text == 'Hey this is Daniel from VoteReminder, wanted to remind you to vote this Tuesday! How do you plan to get to your polling location?',]
length(unique(outbound_messages$phone_number))

#We can use the grep function to see how many people got a message about the environment
#That number is 12,461 so it seems like everyone got it
env_messages = sms_data[grep("environment",sms_data$message_text),]
length(unique(env_messages$phone_number))

```

Combine treatment data with survey data to answer additional questions.

```{r}

#Combine turnout data with treatment group data
df <- merge(rand_data, turnout_data, by = "ai_id")

#Add survey data to dataframe above
df <- merge(df, survey_data, by = "phone_number")

```

###Did the messaging program boost turnout?

In looking at the data, it seems that the treatment group had higher turnout. And the difference is statistically significant.

```{r}

#Create a crosstab of treatment group and turnout
turnout_breakout <- table(df$sms_treat,df$turnout2017)

#Look at crosstab for turnout
#Reorder columns so first column is voters and second column is non-voters
#From first look, it seems like treatment group had higher turnout
turnout_breakout <- turnout_breakout[,c(2,1)]
turnout_breakout

#We must statistically test this by doing a proportions test on our crosstab table
#We find that the Treatment group has stastically higher turnout than the Control group with a p-value of 0.03
prop.test(turnout_breakout, correct=FALSE) 

```


###Did the messaging program convince voters to vote for Jane Smith?

In looking at the data, we find no statistical difference between the treatment and control groups. It seems as the messaging program did not convince voters to vote for Jane Smith.

```{r}
#Create crosstab of treatment group and support for Smith
smith_support_breakout <- table(df$sms_treat,df$support_smith, exclude=NULL)
smith_support_breakout <- smith_support_breakout[,c(2,1)]
smith_support_breakout

#Do a proportion test to see if treatment group has statistically higher rate of support for Smith
#We find that there is no statistical difference in support for Smith between the two groups
prop.test(smith_support_breakout, correct=FALSE) 
```

###Did the message program increase how much voters care about protecting the environment?

The treatment group has an higher average support for the environment, and it is statistically significant.

```{r}

#Use dplyr package to calculate mean, standard deviation, and sample size for environmental protection
#group_by(df[!is.na(df$environment_thermometer),], sms_treat) %>%
#  summarise(mean_value = mean(as.numeric(environment_thermometer)),standard_dev = sd(as.numeric(environment_thermometer)), n_values = n(),
#            std_error = sd(as.numeric(environment_thermometer))/sqrt(n()))

# Mean, SD, N, SE
#1 Control         7.16         2.37     2471     .0477
#2 Treatment       7.33         2.51     2492     .0502

#Use our t-test function to see if the two means are different
#Our function turns a difference of means of .17 with a p-value of 0.015
#We can say that the Treatment group does care more about protecting the environment
t.test2(7.16, 7.33, 2.37, 2.51, 2471, 2492)


```

Graph of turnout by two groups.

```{r}

#Graph for turnout
barplot(c(1358/(1358+2119),1452/(1452+2043)),main="Turnout by Experiment Group", xlab="Group",names.arg=c("Control","Treatment"),
        ylim=c(0,0.5))
text(.65, 0.45, "39.1%")   #1358/(1358+2119)
text(1.75, 0.45, "41.5%")  #1452/(1452+2043)


```

Graph of environmental support by two groups.

```{r}

#Graph for environment thermometer
#create dataframe specifically for this plot
names_vector <- c("Control","Treatment")
means_vector <- c(7.16,7.33)
sd_vector    <- c(2.37,2.51)
n_vector     <- c(2471, 2492)
env_df <- data.frame(names_vector,means_vector,sd_vector, n_vector)
env_df <- do.call(data.frame,env_df)

#Create base plot
env_plot <- barplot(height = env_df$means_vector,
                      beside = true, las = 2,
                      ylim = c(0, 8),
                      cex.names = 0.75, xaxt = "n",
                      main = "Voter Desire to Protect the Environment",
                      ylab = "Mean Environment Protection Score (1-10)",
                      border = "black", axes = TRUE)
#Manually add text labels
text(x = env_plot, y = par("usr")[3]-0.4, srt = 0,
     adj = 1, labels = env_df$names_vector, xpd = TRUE)

#Add confidence bands by manually calculating standard error using standard deviation and sample size
segments(env_plot, env_df$means_vector - env_df$sd_vector/sqrt(env_df$n_vector) * 1.96, env_plot,
         env_df$means_vector + env_df$sd_vector/sqrt(env_df$n_vector * 1.96), lwd = 1.5)
arrows(env_plot, env_df$means_vector - env_df$sd_vector/sqrt(env_df$n_vector) * 1.96, env_plot,
       env_df$means_vector + env_df$sd_vector/sqrt(env_df$n_vector) * 1.96, lwd = 1.5, angle = 90,
       code = 3, length = 0.05)

```